Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2023-02-27 ms/ms03_results.tex
MS

I provide a UCSC track hub with these GWAS colocalizations.
%
In Fig. \ref{fig:1}b-d, I show three screenshots of this UCSC track hub for three different biosamples, namely monocytes, and  frontal cortex biosamples in the chromosome 5 around the gene IL4 (Fig. \ref{fig:1}a).
%
This screenshot shows that depending on the biosample, the regulated genes are very different.
%
In the

@@

These trait class annotations were used to investigate pleiotropy of eQTLs.
%
Different trait classes of eQTLs were aggregated and sorted according to the number of GWAS classes the eQTLs belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The most pleiotropic variants belong to 6 classes that include different traits such as autoimmune diseases, circulatory disases and cancer.
%
The most pleiotropic eQTLs are found in cytobands 1p13.2, 12q24.12 and 19.q13.33.
%
The eQTL rs2476601 is well known for rheumatoid arthritis, juvenile athritis, diabetes and cancer of skin REFs.
%
Our analysis annotates this variant with eQTL genes AP4B1,AP4B1-AS1,BCL2L15,DCLRE1B,MAGI3,PHTF1 and PTPN22 tissues Adipose,Brain,Colon,Fibroblast,ImmuneCell LCL and iPSC

% TODO add rs2522051
This variant rs2522051 is involved in six classes of traits, namely,
allergy, autoimmune diseases, cancer of the breast, circulatory system diseases, hypertension, and respiratory system diseases
(Table \ref{tab:pleitropic_variants}a and Supplementary table 4).
%
The variant rs2522051 is located downstream of the IRF1 gene in cytoband 5q31.1
%
This variant rs2522051 is active in 23 classes of tissues including adipose, brain, digestive cells, immune cells, and sexual organs.
%
This variant rs2522051 regulates several genes such as IL13, PDLIM4, and RAD50.

@@

Pleiotropic eQTLs are concentrated in genomic regions such as in cytobands 3q23,
5q31.1, 9p21.3, 15q24.1, and 19q13.33 for eQTLs involved on five or more classes (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
Therefore, I wanted to compute pleiotropic regions with high concentration of pleiotropic eQTLs.
%
To compute these pleiotropic regions, I use a sliding window to aggregate regions with pleiotropic eQTLs at distances
of less than 1e5 nt (See Methods) (Table \ref{tab:pleiotropic_regions} and Supplementary table 5).
%
% TODO add data
I tested three SNP.H4.PP cutoffs, namely, 0.25, 0.5 and 0.75, to compute these regions.
%
% TODO
For the SNP.PP.H4 cutoff 0.75, we obtain very few regions.
%
A less stringent cutoff 0.5, gives more pleiotropic regions, but still contains the pleiotropic regions found for a
cutoff SNP.PP.H4 0.75.
%
A lower cutoff 0.25 gives more pleiotropic regions, but these regions are different than for the higher cutoffs.
%
Therefore, we fixed our analysis at SNP.PP.H4 0.5.

%
% TODO Verify values
I found seven regions with seven or more trait classes, 30?? regions with 4?? or more trait
classes and 80?? regions with 3?? or more trait classes (Supplementary table 5).
%
% TODO Verify values
50\% of regions with at least two trait classes are shorter than 100 kb, 20\% of regions are
between 100 kb and 200 kb, 10\% between 200 kb and 300 kb, and 20\% are larger than 300 kb (Fig. \ref{fig:pleiotropy_region_distribution}a).

@@

%\subsection*{Do pleiotropic eQTLs have stronger effect sizes?}
%
%Pleiotropic eQTLs affect simultaneously more trait classes.
%%
%Therefore, I wonder about the relationship between pleiotropy, effect size and significance of eQTLs.
%%
%I found that the mean of the absolute eQTL effect sizes (beta) decreased between GWAS class counts 1 and 5 (Fig. \ref{fig:beta_pval}a).
%%
%Regarding the eQTL significance (Negative decimal logarithm of the p-values), I found decreasing mean values for GWAS class counts between 1 and 5 (Fig. \ref{fig:beta_pval}b).
%
%Then I carried out the same analysis for the GWAS effect size (beta) and significance (p-value).
%%
%I found that the mean of the absolute GWAS effect sizes (beta) decreased between GWAS class counts 1 and 5 (Fig. \ref{fig:beta_pval}b).
%%
%Regarding the GWAS variant significance (Negative decimal logarithm of the p-values), I found increasing mean values for GWAS class counts between 1 and 5 (Fig. \ref{fig:beta_pval}b).
%
%These observations suggest that the strength of both eQTL and GWAS effects decrease with the pleiotropy (Fig. \ref{fig:beta_pval}a,b).

>>>>

I provide a UCSC track hub with these GWAS/eQTL colocalizations.
%
In Fig. \ref{fig:1}b-d, I show three screenshots of these UCSC tracks for three different biosamples, namely monocytes, and  frontal cortex biosamples in the chromosome 5 around the gene IL4.
%
In these UCSC tracks, red represents a positive eQTL beta coefficient, which blue represents a negative eQTL beta coefficient.
%
This screenshot shows that depending of gene regulation based on the biosample.
%
In b-cells, there is mainly the SLC22A5 gene that is regulated (Fig. \ref{fig:1}b).
%
On the other hand, monocytes show a significant level changes of P4HA2, SLC22A4 and SLC22A5 (Fig. \ref{fig:1}c).
%
Finally, the cortex biosamples show significant eQTL effects for the PDLIM4, MIR3936HG, IRF-1S1 and RAD50 (Fig. \ref{fig:1}d).

To visualize simultaneously, the GWAS and eQTL beta coefficients, I have create two tracks "beta equal" and "beta unequal".
%
In "beta equal", both GWAS and eQTL beta coefficients have the same sign while in "beta unequal", the coefficients show different sign.
%
This shows that in b-cells, an expression decrease of SLC22A5 gene correlates with a decrease of Asthma predisposition and an increase of extreme height and inflammatory bowel diseases.
%
Altogether, I provide two new resources, namely a web application and a UCSC track hub, that might be useful for the interpretation of genetic human variants.

@@

These trait class annotations were used to investigate pleiotropy of eQTLs.
%
Different trait classes of eQTLs were aggregated and sorted according to the number of GWAS classes the eQTLs belong to (Table \ref{tab:pleitropic_eqtls} and Supplementary table 4).
%
The most pleiotropic variants belong to four classes that include different traits such as autoimmune diseases, circulatory disases and cancer.
%
The most pleiotropic eQTLs are found in the loci around genes ALDH2 (12q24.12), ENO1 (1p36.23), ORMDL3 (17.q21.1), IL4 (5q31.1), BACH2 (6q15), FUT2 (19q13.33), SLC39A8 (4q24) and CYP1A1 (15q24.1)
(Table \ref{tab:pleitropic_eqtls} and Supplementary table 4).
%
We have generated this list based on different SNP.PP.H4 parameter values 0.25, 0.5 and 0.75 and we find the same variants except rs8067378 (Data not shown).
%
Most of these pleiotropic eQTLs are well known with Google Scholar citations (Accessed Feb. 27, 2023).

%By contrast, the variant rs13107325 has 596
% between 596 for rs13107325
%%
%The eQTL rs2476601 is well known for rheumatoid arthritis, juvenile athritis, diabetes and cancer of skin REFs.
%%
%Our analysis annotates this variant with eQTL genes AP4B1,AP4B1-AS1,BCL2L15,DCLRE1B,MAGI3,PHTF1 and PTPN22 tissues Adipose,Brain,Colon,Fibroblast,ImmuneCell LCL and iPSC
%
%% TODO add rs2522051
%This variant rs2522051 is involved in six classes of traits, namely,
%allergy, autoimmune diseases, cancer of the breast, circulatory system diseases, hypertension, and respiratory system diseases
%(Table \ref{tab:pleitropic_eqtls}a and Supplementary table 4).
%%
%The variant rs2522051 is located downstream of the IRF1 gene in cytoband 5q31.1
%%
%This variant rs2522051 is active in 23 classes of tissues including adipose, brain, digestive cells, immune cells, and sexual organs.
%%
%This variant rs2522051 regulates several genes such as IL13, PDLIM4, and RAD50.

@@

Pleiotropic eQTLs are concentrated in genomic regions such as in cytobands 3q23,
5q31.1, 9p21.3, 15q24.1, and 19q13.33 for eQTLs involved on five or more classes (Table \ref{tab:pleitropic_eqtls} and Supplementary table 4).
%
Therefore, I wanted to compute pleiotropic regions with high concentration of pleiotropic eQTLs.
%
To compute these pleiotropic regions, I use a sliding window to aggregate regions with pleiotropic eQTLs at distances
of less than 1e5 nt (See Methods) (Table \ref{tab:pleiotropic_regions} and Supplementary table 5).
%
% TODO add data
I tested three SNP.H4.PP cutoffs, namely, 0.25, 0.5 and 0.75, to compute these regions.
%
% TODO
For the SNP.PP.H4 cutoff 0.75, we obtain very few regions.
%
A less stringent cutoff 0.5, gives more pleiotropic regions, but still contains the pleiotropic regions found for a
cutoff SNP.PP.H4 0.75.
%
A lower cutoff 0.25 gives more pleiotropic regions, but these regions are different than for the higher cutoffs.
%
Therefore, we fixed our analysis at SNP.PP.H4 0.5.

% TODO Verify values
I found seven regions with seven or more trait classes, 30?? regions with 4?? or more trait
classes and 80?? regions with 3?? or more trait classes (Supplementary table 5).
%
% TODO Verify values
50\% of regions with at least two trait classes are shorter than 100 kb, 20\% of regions are
between 100 kb and 200 kb, 10\% between 200 kb and 300 kb, and 20\% are larger than 300 kb (Fig. \ref{fig:pleiotropy_region_distribution}a).

@@

\subsection*{Do pleiotropic eQTLs have stronger effect sizes?}

Pleiotropic eQTLs affect simultaneously more trait classes.
%
Therefore, I wonder about the relationship between pleiotropy, effect size and significance of eQTLs.
%
I found that the mean of the absolute eQTL effect sizes (beta) decreased between GWAS class counts 1 and 5 (Fig. \ref{fig:beta_pval}a).
%
Regarding the eQTL significance (Negative decimal logarithm of the p-values), I found decreasing mean values for GWAS class counts between 1 and 5 (Fig. \ref{fig:beta_pval}b).

Then I carried out the same analysis for the GWAS effect size (beta) and significance (p-value).
%
I found that the mean of the absolute GWAS effect sizes (beta) decreased between GWAS class counts 1 and 5 (Fig. \ref{fig:beta_pval}b).
%
Regarding the GWAS variant significance (Negative decimal logarithm of the p-values), I found increasing mean values for GWAS class counts between 1 and 5 (Fig. \ref{fig:beta_pval}b).

These observations suggest that the strength of both eQTL and GWAS effects decrease with the pleiotropy (Fig. \ref{fig:beta_pval}a,b).

### 2023-02-27 ms/ms08_figures.tex
ENH ucsc tracks
MS

    \begin{subfigure}[]{0.49\textwidth}
        \textbf{a}
        \\
        \includegraphics[width=\textwidth]{\floatRelativePath/plthst_perc_tophits_eqtl.py/hist_perc_tophits_eqtl_excl_mhc.png}
    \end{subfigure}

@@

    \begin{subfigure}[]{.49\textwidth}
        \textbf{c}
        \\
        \includegraphics[width=\textwidth]{\floatRelativePath/pltbar_vep_consequence.py/vep.png}
    \end{subfigure}
    %
    \caption{}
    %
\end{figure}

@@

    \caption{} \label{fig:gwas_egene_etisue_per_variant}
    %
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Fig 9: graphical conclusions
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{figure}[!ht]
    \centering
%
    \begin{subfigure}[]{0.99\textwidth}
        \textbf{a}
        \\
        \includegraphics[width=\textwidth]{fig/model1.png}
    \end{subfigure}

    \caption{\textbf{Model of regulatory variant pleiotropy.} I have investigated three possible mechanisms of pleiotropy. \textbf{Left}, Pleiotropic variants have more eGenes that result in more functions and more phenotypes. This might arise from an enrichment of pleiotropic variants in splicing or 3' UTR regions. \textbf{Center}, I have also found that eGenes of pleiotropic variants are active in more etissues which result in more GWAS phenotypes. This might be explained from variants being bound by more transcription factors. \textbf{Left} Triplets of variant-eGene-eTissues are associated with more GWAS phenotypes, which directly affect the number of GWAS phenotypes. I have found that this might be explained by an enrichment of missense alleles.} \label{fig:beta}
%
\end{figure}

>>>>

    \begin{subfigure}[]{0.32\textwidth}
        \textbf{a}
        \\
        \includegraphics[width=\textwidth]{\floatRelativePath/plthst_perc_tophits_eqtl.py/hist_perc_tophits_eqtl_excl_mhc.png}
    \end{subfigure}

@@

    \begin{subfigure}[]{.49\textwidth}
        \textbf{c}
        \\
        \includegraphics[width=\textwidth]{\floatRelativePath/pltbar_vep_consequence.py/vep.png}
    \end{subfigure}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Fig 9: graphical conclusions
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{figure}[!ht]
    \centering

    \begin{subfigure}[]{0.99\textwidth}
        \textbf{a}
        \\
        \includegraphics[width=\textwidth]{fig/model1.png}
    \end{subfigure}
